--- 
title: "Complex-valued Econometrics with Examples in R"
author: "Sergey Svetunkov and Ivan Svetunkov"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [library.bib, packages.bib, websites.bib]
csl: elsevier-harvard.csl
biblio-style: elsarticle-harv
link-citations: true
github-repo: config-i1/cVAR-book
description: "This monograph introduces new classes of models that use instrument of complex variables. This is for internal use only."
---

# Introduction {-}
The theory of complex variables functions is actively used in a variety of disciplines, including modern physics, engineering sciences [@Mandic2007; @Jia2016], information theory [@Eriksson2006], signal processing [@Arens1957; @Brown1969; @Gardner1993; @Charge2001; @Gerstacker2003; @Delmas2004; @Chevalier2006; @Buzzi2006; @Percival2006; @Taubock2007; @Rubin-Delanchy2007; @Adali2008] and others [e.g. @Kociuba2016]. It is relatively easy to describe the complex phenomena that are studied in these areas of science using the models and methods of this branch of mathematics. Social sciences, being much more complicated due to the unpredictability of human behaviour, tend to use simpler instruments for modelling complex processes, such as linear regression, ARIMA, ETS etc. For example, there are only few papers in which methods and models of the theory of complex variables functions would be used in economics, and they are typically used as instruments for diagnostics or statistical tests [for example, for the unit root test, such as ADF from @Dickey1979] rather than in proper modelling.

In 2012, Springer published the monograph “Complex-valued Modeling in Economics and Finance” [@Svetunkov2012], which presented the theory and methodology of modelling using complex variables in economics. @Svetunkov2012 summarised the main principles of using complex variables in economics and business, and discussed how to estimate some of those models. Before him, the monograph more focused on economics was written by @Tamari1997. The author first introduced Wealth as a complex variable consisting of Output (real part) and Money (imaginary part) and showed how modelling can be done efficiently in the area of economics even with such a simple representation. Unfortunately, this work has gone unnoticed in scientific world. We became aware of this work only in 2016, when Ben Tamari himself kindly sent Sergey Svetunkov his monograph.

Since 2012, there has been some developments in the area of modelling using complex variables in business and economics, notably a paper by @Svetunkov2022 on Complex Exponential Smoothing and @Kourentzes2021a on an error measure based on the idea of complex numbers. Up until now, the modelling with complex variables has been mainly picked up by academics working in the areas of engineering and signal processing. The latter group has been using complex autoregressions for a couple of decades, modelling and predicting signals. In other disciplines, complex variables are not used directly for model building. The probable reason for this is the lack of the communication between the disciplines, poor development of theory of complex-valued modelling and the inherited inertia of academia.

In a try to speed up the adoption of the new instrument, we have written this monograph, summarising the main principles in statistics and econometrics of complex variables. We hope that the theory explained in this monograph will help statisticians, econometricians and forecasters in using complex-valued models for a wider variety of problems.

Chapter 1 discusses the theory of complex numbers, random complex variables, conventional and complex-valued statistics. Chapter 2 moves to the discussion of the simple complex-valued linear regression and how it can be estimated in sample. We then move to Chapter 3 to discuss correlation analysis of complex variables and then extend the material from Chapter 2 to multiple complex linear regression in Chapter 4. After that, we discuss assumptions of complex-valued models and what issues can their violation cause. We finally move to the dynamic complex-valued models in Chapter 6, showing how complex ARIMA can be estimated and what properties it has. Last but not least, we provide two examples of application of complex-valued models, on two datasets, focusing on dynamic and on the regression complex-valued models in Chapter 7. All of this material is supported by examples in R, which will rely heavily on the package `complex`, developed specifically to support this monograph. This means that anyone can then use the proposed models for purposes of time series analysis and forecasting.

<!-- This work was partially supported by the Russian Foundation for Basic Research (grant No. 19-010-00610\19 "Theory, methods and methodologies of forecasting economic development by autoregression models of complex variables"). Thanks to this support, it became possible to carry out this research in general, forming a new scientific direction in statistical modelling and short-term forecasting. -->

<!-- The authors are grateful to those young scientists of St. Petersburg Polytechnic University of Peter the Great, who participated in this study, patiently testing the author's hypotheses. They helped forming the theory behind some of the models discussed in this monograph. We are pleased to mention their names: Evgeny Goltsev, Nikolai Pitukhin, Victoria Matskevich, Yulia Selivanova, Galina Siruk and Nazira Shaikhleeva. -->

This monograph uses the `complex` package for R. Many of the examples in R will use functions from this package, so make sure to install it from CRAN before running any R code [@R-complex]:
```{r eval=FALSE}
install.packages("complex")
```

We will also need several other packages for our analysis, including the following:
```{r eval=FALSE}
library(mvtnorm)
library(tseries)
library(greybox)
```

```{r complexPackage, echo=FALSE, fig.cap="Hex-sticker of the complex package in R."}
knitr::include_graphics("./images/complex.pdf")
```

```{r include=FALSE}
palette("Tableau 10")
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'greybox', 'complex', 'plotrix', 'plotly', 'mvtnorm', 'tseries', 'datasets'
), 'packages.bib')
# Load the necessary packages
library("plotly")
library("latex2exp")
library("kableExtra")
library("statmod")
library("pracma")
library("greybox")
library("plotrix")
library("complex")
library("mvtnorm")
library("tseries")
```


## Who is this monograph for? {-}
This monograph will be useful for PhD students and academics working with multivariate models, such as vector autoregressive models and/or multivariate regression. The models discussed here can in general be applied to many real life problems where a pair of response variables needs to be or can be modelled and/or predicted. 

