# Introduction to theory of complex variables {#intro}
## Theory of complex numbers {#theoryOfComplexNumbers}
This section sets out the basic concepts of the Theory of Complex Variables Functions (TCVF) with some historical overlook of the idea of complex numbers. Hopefully, this will help in understanding the main idea that we will use extensively in the following chapters.

The theory of complex numbers started in 1572 in the small Italian town of Kura. It was in this place and precisely in this year that the manuscript of an Italian mathematician Rafael Bombelli was published. The book was called Algebra. Rafael Bombelli showed in it how to solve the following cubic equation:
\begin{equation}
    x^3 = 15x + 4 .
    (\#eq:BombelliEquation)
\end{equation}
The root of cubic equations at that time was calculated using the formula of Scipione del Ferro. With regards to the task at hand, finding the root should have been carried out as follows:
\begin{equation}
    \begin{aligned}
    x = & \sqrt[3]{\frac{4}{2}+\sqrt{\left(\frac{4}{2}\right)^2 - \left(\frac{15}{3}\right)^3}} + \sqrt[3]{\frac{4}{2}-\sqrt{\left(\frac{4}{2}\right)^2 - \left(\frac{15}{3}\right)^3}} = \\
        & \sqrt[3]{2+\sqrt{-121}} + \sqrt[3]{2-\sqrt{-121}} .
    \end{aligned}
    (\#eq:BombelliSolution)
\end{equation}
As can be seen from equation \@ref(eq:BombelliSolution), there are several square roots, and it follows directly from the right hand side of this equation that to calculate the roots of equation \@ref(eq:BombelliEquation), we need to extract the square root from the negative number, -121. This means that it will not be possible to find a solution to this equation in the domain of real numbers.

This does not mean that equation \@ref(eq:BombelliEquation) does not have a solution. If you represent the problem graphically on a plane then it is easy to see that there is one. But it is impossible to find it arithmetically using the Scipione del Ferro formula, because there is a negative number in the expression on the right-hand side of \@ref(eq:BombelliSolution), and back then, mathematicians were sure that the square root of a negative number does not exist.

The mathematicians of that time struggled to solve this problem, and Rafael Bombelli suggested to ignore the negative sign in the expression (Bombelli, 12). After all, the number -121 can be represented as a product of two numbers: -1 and 121. Then the square root of this number can be written as $\sqrt{-1} \times 11$. So, formula \@ref(eq:BombelliSolution) becomes:
\begin{equation}
    x = \sqrt[3]{2+11\sqrt{-1}} + \sqrt[3]{2-11\sqrt{-1}} .
    (\#eq:BombelliSolutionSqrt)
\end{equation}
Now it is possible to get a solution of the problem, and the root of the cubic equation is $x = 4$.

For a long time, Bombelli's approach was considered by scientists as a convenient mathematical trick. And the square root of minus one was called "magic unit", "vanishing unit", etc. Finally, the term "imaginary unit" was established by René Descartes in 1637 and any number multiplied by the imaginary unit would be called "imaginary number". Merging a real number with an imaginary number gives another number, which can be considered as a more general to the two, which is called "complex". It can be written as:
\begin{equation}
    \underline{z} = x+iy ,
    (\#eq:complexNumber)
\end{equation}
where $x$ is the real part, $iy$ is the imaginary part of a complex number, $x$ and $y$ are real numbers and $i$ is the imaginary unit that satisfies the equality:
\begin{equation}
    i^2 = -1 .
    (\#eq:imaginaryUnit)
\end{equation}
We use the symbol $\underline{}$ to denote a complex number.

With complex numbers, one can perform almost all the same operations as with the real ones. However, taking into account the properties of the imaginary unit, these operations can lead to results that are not common in the domain of real numbers.

The main problem that analysts face when trying to understand a complex number is the complexity of the interpretation of the imaginary part. The main question in this case can be formulated as: Where do imaginary numbers appear in practice? And what is the practical meaning of an imaginary unit? We have heard questions like that many times, and while they are valid, we think that they do not have appropriate answers and divert the analysts from what the modelling with complex variables brings.

We argue that the imaginary unit does not have any practical meaning: neither in economics, nor in sociology, nor in engineering, nor in physics. The imaginary unit is a mathematical rule, and that is all. It is a mathematical abstract that has useful properties. A number $\sqrt{2}$ or $\ln 3$ do not have practical meaning either, but they are useful for modelling purposes and solving applied problems in a variety of disciplines. Similarly, an imaginary unit is convenient for solving a whole class of problems in different areas of application. With the help of the rules specified by conditions \@ref(eq:complexNumber) and \@ref(eq:imaginaryUnit), it is possible to use new mathematical operations, obtain new mathematical results, and create new mathematical models.

Imaginary and complex numbers are mathematical instruments that can help in describing a real-life phenomenon. If a researcher decides to use complex variables for real processes modelling, they will need to predefine the rules according to which one component of a complex process is attributed to the real part, and another one is attributed to the imaginary part of the complex variable. These rules might seem arbitrary, because there is no "true" recipe, according to which the assignment to complex variable parts should be done. The main motivation for them is the efficiency of modelling (such as goodness of fit or forecasting accuracy) and nothing else.

In order to better understand how the modelling with complex variables can be done, we need to better understand how complex variables can be represented and interpreted. A real number represents visually a certain segment on the numerical axis, which has a zero point and a multitude of values from minus infinity to infinity. Any real number is characterized by the distance from zero to the value of that number. Negative numbers are located to the left of zero, while the positive ones are located to the right of it.

A complex number consists of two parts, which can be visualised on the plane with two perpendicular axes, where the real numbers are marked on the horizontal line, and the imaginary ones are on the vertical one. This is shown visually in Figure \@ref(fig:complexPlane) with a complex number $\underline{z}=x+iy$.

```{r complexPlane, fig.width=3, fig.height=2, fig.cap="Visual presentation of a complex number on the complex plane.", echo=FALSE}
par(mar=c(1,1,1,1), cex=1.25, cex.lab=1/1.25)
plot(c(0,9),c(0,6), type="l",
     xlab="", ylab="",
     xlim=c(0,10), ylim=c(0,8), axes=F,
     yaxt="n", xaxt="n", col="white")
lines(c(0,9), c(6,6), lwd=2, lty=2, col="grey")
lines(c(9,9), c(0,6), lwd=2, lty=2, col="grey")
arrows(0,0,9,6, lwd=2, angle=15)
text(9,6,TeX("$\\underline{z}"),pos=3)
text(9.3,0,"x",pos=3)
text(0,6.4,"y",pos=4)
abline(h=0,v=0,lty=1)
title(ylab="Imaginary number", line=0)
title(xlab="Real number", line=0)
```

Any point lying on the complex plane in Figure \@ref(fig:complexPlane) characterises a complex number, even if that point lies on the axis of real numbers. In that specific case, we would be talking about a complex number with a zero imaginary part, i.e. $\underline{z}=x+i0$.

Given that complex numbers can be represented on a plane (in so called Cartesian coordinates), the number \@ref(eq:complexNumber) can be represented in a form of a vector that starts in the origin of coordinates and ends at the point (x, y). Then any complex number can be represented in form of polar coordinates using the magnitude of the vector and its polar angle:
\begin{equation}
    \underline{z} = x+iy = r (\cos \phi + i \sin \phi),
    (\#eq:complexNumberTrigonometric)
\end{equation}
where $r=|\underline{z}|$ is the magnitude, which is calculated as a Euclidean distance from the origin to the point (x, y) on the plane:
\begin{equation}
    r = \sqrt{x^2 + y^2},
    (\#eq:complexNumberMagnitude)
\end{equation}
and $\phi=\Arg(\underline{z})$ is the angle, which equals to:
\begin{equation}
    \phi = \arctan \frac{y}{x} + 2 \pi l,
    (\#eq:complexNumberAngle)
\end{equation}
where $l$ is an integer number. While depending on a task, $l$ can be set to a positive, a negative number, or a zero, for convenience, we will restrict it to $l=0$, because all the other values will not be useful for the inference in following chapters. The polar angle is sometimes referred to as the argument of a complex number.

Using the magnitude and the polar angle, we can also represent any complex number in the exponential form, which was first proposed in 1748 by Euler, in his book "Introduction to the Infinitesimal Analysis" **(Euler, 1961, pp. 118-119)**, which:
\begin{equation}
    \underline{z} = r e^{i \phi} ,
    (\#eq:complexNumberExponential)
\end{equation}
where $e$ is the Euler's constant. Equation \@ref(eq:complexNumberExponential) is nowadays also called "Euler's form". Comparing \@ref{eq:complexNumberTrigonometric} with \@ref{eq:complexNumberExponential}, we can conclude that:
\begin{equation}
    e^{i \phi} = \cos \phi + i \sin \phi ,
    (\#eq:EulerFormula)
\end{equation}
which gives the connection between the linear, trigonometric and exponential forms of a complex number. In the exponential form \@ref(eq:complexNumberExponential), a positive real number has $\phi=0$, while a negative one has $\phi=\pi$, and all imaginary numbers have an angle dividable by $\frac{\pi}{2}$. So, for example:
\begin{equation*}
    \begin{aligned}
    41 = & 41 e^{i 0} \\
    i41 = & 41 e^{i \frac{\pi}{2}} \\
    -41 = & 41 e^{i \pi} \\
    -i41 = & 41 e^{i \frac{3 \pi}{2}}
    \end{aligned}
\end{equation*}
In fact, multiplication of any complex number by the imaginary unit implies the rotation of complex vector by $\frac{\pi}{2}$, as shown in Figure \@ref(fig:complexPlaneMultiplication), where a complex number $z_1 = x_1 + i y_1$ becomes $z_2 = z_1 \times i = x_2 + i y_2$ etc.

```{r complexPlaneMultiplication, fig.width=3, fig.height=3, fig.cap="Multiplication of a complex number by $i$ implies the rotation of it by $\\frac{\\pi}{2}$.", echo=FALSE}
x <- complex(real=9, imaginary=6)
par(mar=c(1,1,1,1), cex=1)
plot(c(0,9),c(0,6), type="l",
     xlab="", ylab="",
     xlim=c(-12,15), ylim=c(-12,15), axes=F,
     yaxt="n", xaxt="n", col="white")
draw.circle(0,0,sqrt(9^2+6^2))
draw.arc(0,0,2,Arg(x),Arg(x)+pi/2, lwd=2, col=2)
text(Re(x),Im(x),TeX("$z_1$", italic=T),pos=3)
arrows(0,0,Re(x),Im(x), lwd=2, angle=15, col=1)
text(1,1.5,TeX("$\\frac{\\pi}{2}$", italic=T),pos=3, cex=0.8)
x <- x * 1i
draw.arc(0,0,2,Arg(x),Arg(x)+pi/2, lwd=2, col=3)
text(Re(x),Im(x),TeX("$z_2$", italic=T),pos=3)
arrows(0,0,Re(x),Im(x), lwd=2, angle=15, col=2)
x <- x * 1i
draw.arc(0,0,2,Arg(x),Arg(x)+pi/2, lwd=2, col=4)
text(Re(x),Im(x),TeX("$z_3$", italic=T),pos=1)
arrows(0,0,Re(x),Im(x), lwd=2, angle=15, col=3)
x <- x * 1i
draw.arc(0,0,2,Arg(x),Arg(x)+pi/2, lwd=2, col=1)
text(Re(x),Im(x),TeX("$z_4$", italic=T),pos=1)
arrows(0,0,Re(x),Im(x), lwd=2, angle=15, col=4)
abline(h=0,v=0,lty=1)
text(15, -1, "x", cex=1.25)
text(-1, 15, "y", cex=1.25)
```

In a way, any mathematical operation with a complex number implies a change in magnitude and angle of the number, while the multiplication and division by a number with non-zero imaginary part will always lead to the rotation of the vector. Operations of addition and subtraction are simpler in the linear form:
\begin{equation*}
    z_3 = z_1 + z_2 = x_1 + x_2 + i (y_1 + y_2) ,
\end{equation*}
while operations of multiplication and division are easier to do in either exponential or trigonometric forms of complex numbers:
\begin{equation*}
    z_3 = z_1 \times z_2 = r_1 r_2 e^{i \phi_1 + \phi_2} 
\end{equation*}
or:
\begin{equation*}
    z_3 = r_1 r_2 \left(\cos (\phi_1 + \phi_2) + i \sin (\phi_1 + \phi_2) \right) .
\end{equation*}

Furthermore, given that any complex number can be represented in the exponential form \@ref(eq:complexNumberExponential), it can also be visualised on a polar coordinates plane, where the magnitude is marked on the x-axis, and the polar angle is marked on the y-axis. This is shown visually in Figure \@ref(fig:complexPlanePolar).

```{r complexPlanePolar, fig.width=3, fig.height=2, fig.cap="Visual presentation of a complex number in the polar coordinates.", echo=FALSE}
par(mar=c(1,1,1,1), cex=1.25, cex.lab=1/1.25)
plot(0,0, type="p",
     xlab="", ylab="",
     xlim=c(0,12), ylim=c(0,2), axes=F,
     yaxt="n", xaxt="n", col="white")
lines(c(0,abs(9+6i)), c(Arg(9+6i),Arg(9+6i)), lwd=2, lty=2, col="grey")
lines(c(abs(9+6i),abs(9+6i)), c(0,Arg(9+6i)), lwd=2, lty=2, col="grey")
points(abs(9+6i),Arg(9+6i), lwd=2, pch=20)
text(abs(9+6i),Arg(9+6i),TeX("$\\underline{z}$"),pos=3)
text(abs(9+6i)*(1+0.05),0,"r",pos=3)
text(0,Arg(9+6i)*(1+0.15),TeX("$\\phi$", italic=T),pos=4)
abline(h=0,v=0,lty=1)
title(ylab="Polar angle", line=0)
title(xlab="Magnitude", line=0)
```

The usefulness of the polar coordinates presentation becomes apparent when a set of complex numbers is considered, because then in some cases it becomes possible to see some relations that are not obvious on the Cartesian plane. Figure \@ref(fig:complexCartesianvsPolar) shows an example of a set of complex random numbers, for which the real and imaginary parts do not seem to have any obvious linear relation, but the magnitude and the angle have a negative relation.

```{r complexCartesianvsPolar, fig.height=4, fig.cap="Visualisation of a set of complex numbers on Cartesian and on polar coordinates planes.", echo=FALSE}
x <- rnorm(100,100,10)
y <- 1/x*rlnorm(100,0,0.02)
z <- x*exp(y*1i)
par(mar=c(1,1,1,1), cex=1.25, mfcol=c(1,2))
plot(Re(z), Im(z), type="p",
     xlab="", ylab="",
     yaxt="n", xaxt="n")
title(ylab="Imaginary number", line=0)
title(xlab="Real number", line=0)

plot(abs(z),Arg(z), type="p",
     xlab="", ylab="",
     yaxt="n", xaxt="n")
points(abs(9+6i),Arg(9+6i), lwd=2, pch=20)
abline(h=0,v=0,lty=1)
title(ylab="Polar angle", line=0)
title(xlab="Magnitude", line=0)
```

In this case, modelling can be done in the exponential form of complex numbers, which might allow capturing complicated non-linear relations between variables.

When it comes to comparing complex numbers, mathematically we will say that two complex numbers $z_1 = x_1 + i y_1$ and $z_2 = x_2 + i y_2$ are equal to each other if and only if their real and imaginary parts are equal:
\begin{equation*}
    z_1 = z_2 \iff \left \lbrace
    \begin{aligned}
        & x_1 = x_2 \\
        & y_1 = y_2
    \end{aligned}
    \right. ,
\end{equation*}
which is equivalent to saying that their magnitudes and polar angles are equal:
\begin{equation*}
    z_1 = z_2 \iff \left \lbrace
    \begin{aligned}
        & |z_1| = |z_2| \\
        & \Arg(z_1) = \Arg(z_2)
    \end{aligned}
    \right. .
\end{equation*}

Unfortunately, given that complex numbers are two dimensional, it is not possible to say in general whether one number is greater or less than the other. However, we could use magnitude to compare complex numbers, to say which one lies further away from the origin than the other. In that case, the number with a larger magnitude could be considered as a greater than the one with the smaller magnitude. While we could compare the polar angles as well, they only give us information about rotation of a complex number and thus do not give any meaningful information about the comparison of numbers. So, two complex numbers in Figure \@ref(fig:complexPlaneCircle) would have the same magnitude, but will have different angles, and it is not possible to say whether one number is greater than the other in this case. In order to conclude that, we would need to devise additional criteria for comparison of numbers (e.g. positive numbers are "better" than the negative ones, thus $\phi_1=\pi$ and $\phi_1=-\pi$ is "worse" than $\phi_2=0$).

```{r complexPlaneCircle, fig.width=4, fig.height=4, fig.cap="Comparison of two complex numbers $z_1$ and $z_2$ that have the same magnitude $r$, but different angles $\\phi_1$ and $\\phi_2$.", echo=FALSE}
par(mar=c(1,1,1,1), cex=1.25, cex.lab=1/1.25)
plot(c(0,9),c(0,6), type="l",
     xlab="", ylab="",
     xlim=c(0,12), ylim=c(0,12), axes=F,
     yaxt="n", xaxt="n", col="white")
draw.circle(0,0,sqrt(9^2+6^2))
text(9,6,TeX("$z_1$", italic=T),pos=3)
text(6,9,TeX("$z_2$", italic=T),pos=3)
text(4.5,3,TeX("$r$", italic=T),pos=3)
text(3,4.5,TeX("$r$", italic=T),pos=3)
draw.arc(0,0,2,0,atan(6/9), col="grey", lwd=2)
draw.arc(0,0,4,0,atan(9/6), col="grey", lwd=2)
arrows(0,0,9,6, lwd=2, angle=15)
arrows(0,0,6,9, lwd=2, angle=15)
text(2.7,0.1,TeX("$\\phi_1$", italic=T),pos=3)
text(4.5,1,TeX("$\\phi_2$", italic=T),pos=3)
abline(h=0,v=0,lty=1)
title(ylab="Imaginary number", line=0)
title(xlab="Real number", line=0)
```

Continuing the discussion of different forms of a complex number, its magnitude can be represented in exponential form:
\begin{equation*}
    r = e^{\ln(r)},
\end{equation*}
where $\ln$ is the natural logarithm. This can then be used to rewrite the exponential form of a complex number as:
\begin{equation}
    \underline{z} = e^{\ln r + i \phi} ,
    (\#eq:complexNumberExponentialAll)
\end{equation}
And this form allows calculating logarithms of a complex number:
\begin{equation*}
    \ln \underline{z} = \ln \left(e^{\ln r + i \phi} \right) = \ln r + i \phi .
\end{equation*}
This operation will become useful when we discuss transformations of complex-valued functions, but it also shows that a logarithm of a negative real number is a complex number, because in that case $\phi=\pi$. This demonstrates that the field of complex numbers is complete: any mathematical operation of a complex number will give another complex number.

One of the important definitions in theory of complex numbers is the conjugate complex number. It is the number, for which the imaginary part has the sign opposite to the one of the original variable. For example, the conjugate of the complex variable $\underline{z} = x+ iy$ is $\underline{\tilde{z}} = x- iy$. These numbers are useful because the multiplication of a complex number by its conjugate gives a real number:
\begin{equation}
    \underline{z} \times \underline{\tilde{z}} = (x+ iy) (x- iy) = x^2 + y^2 ,
    (\#eq:complexNumberConjugateMulti)
\end{equation}
which is the square of the magnitude \@ref(eq:complexNumberMagnitude) of both complex variables $\underline{z}$ and $\underline{\tilde{z}}$. Conjugation is distributive over addition, subtraction, multiplication and division, meaning that:
\begin{equation}
    \begin{aligned}
        & \tilde{x + y} = \tilde{x} + \tilde{y} \\
        & \tilde{x - y} = \tilde{x} - \tilde{y} \\
        & \tilde{x \times y} = \tilde{x} \times \tilde{y} \\
        & \tilde{\frac{x}{y}} = \frac{\tilde{x}}{\tilde{y}} .
    \end{aligned}
    (\#eq:complexNumberConjugateDistributive)
\end{equation}

Finally, given that any complex number can be represented visually in a form of a vector, we can write it mathematically either like vector:
\begin{equation}
    \begin{pmatrix} x \\ y \end{pmatrix}
    (\#eq:complexNumberVectors)
\end{equation}
or like a matrix:
\begin{equation}
    \begin{pmatrix} x & -y \\ y & x \end{pmatrix} .
    (\#eq:complexNumberMatrix)
\end{equation}
In the latter case, the real number can be represented as a diagonal matrix, while the imaginary one is the matrix with zero diagonal:
\begin{equation*}
    \begin{aligned}
        & \begin{pmatrix} x & 0 \\ 0 & x \end{pmatrix} \\
        & \begin{pmatrix} 0 & -y \\ y & 0 \end{pmatrix}
    \end{aligned} .
\end{equation*}
All the mathematical operations done with the vector and matrix representations would correspond to the ones for the conventional complex numbers. Furthermore, the multiplication by the transpose of the original object in both of these cases is equivalent to the multiplication by the conjugate value \@ref(eq:complexNumberConjugateMulti). For the vector:
\begin{equation}
    \begin{pmatrix} x & y \end{pmatrix} \times \begin{pmatrix} x \\ y \end{pmatrix} = \begin{pmatrix} x^2 + y^2 \end{pmatrix}
    (\#eq:complexNumberVectorsMulti)
\end{equation}
and for the matrix:
\begin{equation}
    \begin{pmatrix} x & -y \\ y & x \end{pmatrix} \times \begin{pmatrix} x & y \\ -y & x \end{pmatrix} = 
    \begin{pmatrix} x^2 + y^2 & 0 \\ 0 & x^2 + y^2 \end{pmatrix}.
    (\#eq:complexNumberMatrixMulti)
\end{equation}
These transpositions will be called in this monograph "conjugate transpositions". The vector and matrix representations become especially useful when we work with complex variables to construct complex models. Finally, in case of matrix form, it is possible to multiply complex variable by itself without the conjugation, which results in:
\begin{equation}
    \begin{pmatrix} x & -y \\ y & x \end{pmatrix} \times \begin{pmatrix} x & -y \\ y & x \end{pmatrix} = 
    \begin{pmatrix} x^2 - y^2 & - 2 xy \\ 2 xy & x^2 - y^2 \end{pmatrix}.
    (\#eq:complexNumberMatrixMultiDirect)
\end{equation}
In this monograph, we will call the multiplications \@ref(eq:complexNumberMatrixMulti) and \@ref(eq:complexNumberMatrixMultiDirect)  "conjugate" and "direct" respectively.


### Complex variables {#complexVariable}
Similarly to any other variable, complex variable is a place holder for values, with the difference that it represents two parts of a number instead of one. Given the properties of complex numbers discussed in the previous section, any complex variable can be represented in linear, trigonometric and exponential forms. But in addition, it can also be represented in the form of a set of equations, i.e. if $y_r + i y_i = 3+2i$ then:
\begin{equation}
    \begin{aligned}
        y_r = & 3 \\
        y_i = & 2
    \end{aligned} ,
    (\#eq:complexVariablesSystem)
\end{equation}
where the subscript $r$ or $i$ refers to respectively the real or imaginary parts of the complex number. In fact, any function of a complex variable can be represented as system of two equations, so that for a function
\begin{equation}
    y_{r} + i y_{i} = f(x_r+ix_i)
    (\#eq:complexFunction)
\end{equation}
we have:
\begin{equation}
    \begin{aligned}
        y_r = & \mathcal{R}(f(x_r+ix_i)) \\
        y_i = & \mathcal{I}(f(x_r+ix_i))
    \end{aligned} ,
    (\#eq:complexFunctionSystem)
\end{equation}
where $\mathcal{R}()$ and $\mathcal{I}()$ are respectively the real and the imaginary parts of a resulting complex variable. For example, a simple linear function:
\begin{equation*}
    y_r + i y_i = a_{0,r} + i a_{0,i} + (a_{1,r} + i a_{1,i}) (x_{1,r} + i x_{1,i}),
\end{equation*}
which can be written after opening the brackets and regrouping elements as:
\begin{equation*}
    y_r + i y_i = a_{0,r} + a_{1,r} x_{1,r} - a_{1,i} x_{1,i} + i \left( a_{0,i} + a_{1,i} x_{1,r} + a_{1,r} x_{1,i} \right)
\end{equation*}
and is equivalent to the system of the following two equations:
\begin{equation*}
    \begin{aligned}
        y_r = & a_{0,r} + a_{1,r} x_{1,r} - a_{1,i} x_{1,i} \\
        y_i = & a_{0,i} + a_{1,i} x_{1,r} + a_{1,r} x_{1,i}
    \end{aligned} .
\end{equation*}
Given the vector and matrix representation of complex variables, the same linear function can be written as:
\begin{equation*}
    \begin{pmatrix} y_r \\ y_i \end{pmatrix} = \begin{pmatrix} a_{0,r} \\ a_{0,i} \end{pmatrix} + \begin{pmatrix} a_{1,r} & -a_{1,i} \\ a_{1,i} & a_{1,r} \end{pmatrix} \begin{pmatrix} x_{1,r} \\ x_{1,i} \end{pmatrix} ,
\end{equation*}
which is convenient in some situations. Presenting the system of equations \@ref(eq:complexFunctionSystem) in the linear form \@ref(eq:complexFunction) is often convenient. In fact, any function of complex variables can be represented as a system of two functions. Even if the real or the imaginary part of the output complex variable $y_r + i y_i$ is equal to zero, we can still use a system of two equations, where one of the equations becomes just a constraint for the function. For example, if we assume that $y_i=0$, then the system \@ref(eq:complexFunctionSystem) becomes:
\begin{equation*}
    \begin{aligned}
        & y_r = \mathcal{R}(f(x_r+ix_i)) \\
        & \mathcal{I}(f(x_r+ix_i)) = 0
    \end{aligned} .
\end{equation*}
This property becomes especially useful if one needs to construct a model with linear relations between the input variables [this is discussed in Chapter ??? of @Svetunkov2014].

Given the discussion about the vector and matrix forms of complex numbers, the system of equations \@ref(eq:complexFunctionSystem) can be written as:
\begin{equation}
    \begin{pmatrix} y_r \\ y_i \end{pmatrix} =
    \begin{pmatrix} \mathcal{R}(f(x_r+ix_i)) \\ \mathcal{I}(f(x_r+ix_i)) \end{pmatrix} .
    (\#eq:complexFunctionVectors)
\end{equation}
or:
\begin{equation}
    \begin{pmatrix} \mathcal{R}(f(x_r+ix_i)) & - \mathcal{I}(f(x_r+ix_i)) \\ \mathcal{I}(f(x_r+ix_i)) & \mathcal{R}(f(x_r+ix_i)) \end{pmatrix} .
    (\#eq:complexFunctionMatrix)
\end{equation}
The representations \@ref(eq:complexFunction), \@ref(eq:complexFunctionSystem), \@ref(eq:complexFunctionVectors) and \@ref(eq:complexFunctionMatrix) are equivalent and can be used interchangeably depending on the task at hand.

Finally, in terms of definitions, the function \@ref(eq:complexFunction) is often referred to as a "complex-valued" function. It maps a certain values of the input variable $x_r + i x_i$ with a specific value of the output variable $y_r +i y_i$. If there is only a one-to-one relation between the input and output variables then such function is called "univalent". When several different values of the input variable leads to one and the same value of the output variable then the function is called "multivalent". An example of the former is the linear function:
\begin{equation*}
    y_r + i y_i = (a_{1,r} + i a_{1,i}) (x_{1,r} + i x_{1,i}),
\end{equation*}
while for the latter, an example is
\begin{equation*}
    y_r + i y_i = (a_{1,r} + i a_{1,i}) (x_{1,r} + i x_{1,i})^2, 
\end{equation*}
where two different input complex numbers (for example $0 + i$ and $0-i$) correspond to the same output number ($-1 + i0$). While there are examples of functions like that in the real domain, in the complex one, they appear even more often due to the rotation property discussed in the previous Subsection.


### Vectors of complex variables {#vectorComplexVariables}
Last but not least, for what follows, we introduce vectors and matrices of complex variables. There are different ways how one can represent a vector of complex variables. The simplest and most direct one is:
\begin{equation}
    \underline{\mathbf{x}} = \begin{pmatrix} x_{r,1} + i x_{i,1} \\ x_{r,2} + i x_{i,2} \\ x_{r,3} + i x_{i,3} \end{pmatrix} ,
    (\#eq:complexVector)
\end{equation}
where the index in subscript (1, 2, 3) represents the observed first, second and third values of the complex variable.  Alternatively, given that any complex variable can be represented as a vector \@ref(eq:complexNumberVectors), the same complex vector can be written as a matrix:
\begin{equation}
    \mathbf{x} = \begin{pmatrix} x_{r,1} & x_{i,1} \\ x_{r,2} & x_{i,2} \\ x_{r,3} & x_{i,3} \end{pmatrix} ,
    (\#eq:complexVectorMatrix)
\end{equation}
Finally, given the connection of complex numbers and matrices \@ref(eq:complexNumberMatrix), the same complex vector $x$ can be written as a matrix:
\begin{equation}
    \underset{\sim}{\mathbf{x}} = \begin{pmatrix} x_{r,1} & - x_{i,1} \\ x_{i,1} & x_{r,1} \\
                                 x_{r,2} & - x_{i,2} \\ x_{i,2} & x_{r,2} \\
                                 x_{r,3} & - x_{i,3} \\ x_{i,3} & x_{r,3} \end{pmatrix} .
    (\#eq:complexMatrix)
\end{equation}
We use the symbol $\sim$ below a character to denote the complex variable in the matrix form \@ref(eq:complexMatrix). Each of the forms \@ref(eq:complexVector), \@ref(eq:complexVectorMatrix) and \@ref(eq:complexMatrix) might be useful in different circumstances, their usage should be dictated by the modelling purpose.

Finally, note that the transposition of these three forms will imply different things. As discussed earlier, in vector and matrix forms transposition of a complex number implies the conjugation. However, transposing \@ref(eq:complexVector) will not have the same effect - it will only switch rows and columns of the complex vector:
\begin{equation}
    \underline{\mathbf{x}}^{\top} = \begin{pmatrix} x_{r,1} + i x_{i,1} & x_{r,2} + i x_{i,2} & x_{r,3} + i x_{i,3}
                        \end{pmatrix} ,
    (\#eq:complexVectorTransposed)
\end{equation}
If we then switch to either the form \@ref(eq:complexNumberVectors) or \@ref(eq:complexNumberMatrix), we will not get the vector of conjugate complex numbers, but rather an object of the original complex variables with the switched rows and columns:
\begin{equation}
    {\mathbf{x}}^{\top} = \begin{pmatrix} x_{r,1} & x_{r,2} & x_{r,3} \\
                                 x_{i,1} & x_{i,2} & x_{i,3}
                        \end{pmatrix} 
    (\#eq:complexVectorMatrixTransposed)
\end{equation}
and
\begin{equation}
    \underset{\sim}{\mathbf{x}}^{\top} = \begin{pmatrix} x_{r,1} & - x_{i,1} & x_{r,2} & - x_{i,2} & x_{r,3} & - x_{i,3} \\
                                        x_{i,1} & x_{r,1} & x_{i,2} & x_{r,2} & x_{i,3} & x_{r,3}
                        \end{pmatrix} .
    (\#eq:complexMatrixTransposed)
\end{equation}
We will denote the operation of transposition above with the symbol $\top$ and will call it just "transposition". The transposition that produces the conjugate complex numbers is called "conjugate transposition" and will be denoted as $\prime$ in this monograph. For the three examples above, the conjugate transposition will be:
\begin{equation}
    \underline{\mathbf{x}}^{\prime} = \begin{pmatrix} x_{r,1} - i x_{i,1} & x_{r,2} - i x_{i,2} & x_{r,3} - i x_{i,3}
                        \end{pmatrix} ,
    (\#eq:complexVectorTransposedConj)
\end{equation}
\begin{equation}
    {\mathbf{x}}^{\prime} = \begin{pmatrix} x_{r,1} & x_{r,2} & x_{r,3} \\
                                         -x_{i,1} & -x_{i,2} & -x_{i,3}
                        \end{pmatrix} 
    (\#eq:complexVectorMatrixTransposedConj)
\end{equation}
and
\begin{equation}
    \underset{\sim}{\mathbf{x}}^{\prime} = \begin{pmatrix} x_{r,1} & x_{i,1} & x_{r,2} & x_{i,2} & x_{r,3} & x_{i,3} \\
                                        -x_{i,1} & x_{r,1} & -x_{i,2} & x_{r,2} & -x_{i,3} & x_{r,3}
                        \end{pmatrix} .
    (\#eq:complexMatrixTransposedConj)
\end{equation}


## Complex random variable {#complexRandomVariable}
In statistics, a random variable is a variable, the value of which depends on random events. For example, for the classical coin tossing experiment, $x$ would be considered as a random variable if it is equal to 1 in case of heads and is equal to zero otherwise. In case of complex random variables (c.r.v.), the logic is similar, but we would typically deal with two-dimensional cases (e.g. tossing two coins simultaneously). We accept that in some cases, some parts of the c.r.v. might become non-random (e.g. the real part becomes equal to some fixed number). As such, we would be saying that $x=x_r+ix_i$ is a complex random variable if at least one part of it follows some distribution. Furthermore, while random variables can be measured in a variety of scales (coin tossing represents the nominal scale), in this monograph we focus the discussion on continuous numerical variables.

As any other random variable, c.r.v. can be characterised by its moments. However, some of its moments differ from the conventional ones for real variables and are based on the properties of complex numbers.


### First moment
The first moment of any c.r.v. is relatively straightforward. For the variable $\underline{x}=x_r+ix_i$ it is:
\begin{equation}
    \underline{\mu} = \mathrm{E}(\underline{x}) = \mathrm{E}(x_r + i x_i) = \mathrm{E}(x_r) + i \mathrm{E}(x_i) = \mu_{r} + i \mu_{i},
    (\#eq:crvMomentFirst)
\end{equation}
due to the law of total expectation and because $i$ is a constant. In \@ref(eq:crvMomentFirst), $\mathrm{E}(.)$ is the expectation of a variable, $\underline{\mu}$ is the first moment of a complex variable and $\mu_{r}$ and $\mu_{i}$ are the respective real and imaginary parts of the first moment. In sample, this moment can be calculated as:
\begin{equation}
    \underline{\hat{\mu}} = \hat{\mu}_{r} + i \hat{\mu}_{i} = \frac{1}{n}\sum_{j=1}^n x_{r,j} + i \frac{1}{n}\sum_{j=1}^n x_{i,j},
    (\#eq:crvMomentFirstSample)
\end{equation}
where $n$ is the sample size and $\hat{\mu}$ is the estimate of $\mu$. The properties of the first moment of c.r.v. are similar to the properties of any other random variable. The only thing to note is that the moment can also be represented in a vector or matrix form (see Subsection \@ref(complexVariable)):
\begin{equation}
    \begin{pmatrix} \mu_{r} \\ \mu_{i} \end{pmatrix} 
    (\#eq:crvMomentsVector)
\end{equation}
or
\begin{equation}
    \begin{pmatrix} \mu_{r} & - \mu_{i} \\ \mu_{i} & \mu_{r} \end{pmatrix} .
    (\#eq:crvMomentsMatrix)
\end{equation}

In R, the first moment can be obtained via the `mean()` function from `stats` package:

```{r}
# Create a random variable
x <- complex(real=rnorm(100, mean=50, sd=10),
             imaginary=rnorm(100, mean=100, sd=10))
# Calculate the first moment
mean(x)
```


### Second moment {#crvSecondMoment}
While there are raw moments for c.r.v., here we will focus on the centred ones, i.e. moments for $x-\bar{x}$. There are several instruments related to the second moment that could characterise a complex random variable. The first one is called "variance" and is based on the multiplication by a conjugate number [@reference]. So for the c.r.v. $x_r + i x_i$ the variance is:
\begin{equation}
    \begin{aligned}
    \sigma_x^2 = & \mathrm{E}((x-\mu) (\tilde{x}-\tilde{\mu})) = \\
                 & \mathrm{E}\left(((x_r-\mu_{r}) + i (x_i-\mu_{i}))((x_r-\mu_{r}) - i (x_i-\mu_{i}))\right) = \\
                 & \mathrm{E}((x_r-\mu_{r})^2) +  \mathrm{E}((x_i-\mu_{i})^2)
    \end{aligned}
    (\#eq:crvMomentSecondVariance)
\end{equation}
or taking that the variances of the real and the imaginary parts are $\sigma_{x_r}^2$ and $\sigma_{x_i}^2$ respectively:
\begin{equation}
    \sigma_x^2 = \sigma_{x_r}^2 + \sigma_{x_i}^2.
    (\#eq:crvMomentSecondVarianceShort)
\end{equation}
Multiplication by conjugate in \@ref(eq:crvMomentSecondVariance) allows keeping the size of variability for both real and imaginary parts, but at the same time masks the individual contribution of each part in the overall variance and removes the covariance between the parts. The resulting values shows in a way the overall variability of a c.r.v. along a diagonal line as shown in Figure \@ref(fig:crvMomentSecondVariance).

```{r crvMomentSecondVariance, fig.height=4, fig.width=4, fig.cap="Visual representation of variance of a complex random variable.", echo=FALSE}
xr <- rnorm(1000,100,10)
xi <- xr + rnorm(1000,0,10)
x <- complex(real=xr, imaginary=xi)
plot(x, col="lightblue", xlim=c(70, 130), ylim=c(70,130))
abline(a=-40-sqrt(2), b=sqrt(2), col="black", lwd=2, lty=2)
lines(100+c(-10,10)/2, rep(100-10*sqrt(2)/2,2), col="black", lwd=3)
lines(rep(100+10/2,2), 100+c(-10,10)*sqrt(2)/2, col="black", lwd=3)
lines(100+c(-10,10)/2, 100+c(-10,10)*sqrt(2)/2, col="black", lwd=3)
text(100, 90, latex2exp::TeX("$\\sigma_{x_r}$", italic=T))
text(108, 100, latex2exp::TeX("$\\sigma_{x_i}$", italic=T))
text(100, 100, latex2exp::TeX("$\\sigma_{x}$", italic=T), pos=2)
```

Figure \@ref(fig:crvMomentSecondVariance) demonstrates graphically the relation between standard deviations of real, imaginary and the overall variance of a complex variable according to the formula \@ref(eq:crvMomentSecondVarianceShort). 

@Picinbono note that the moment \@ref(eq:crvMomentSecondVarianceShort) is not sufficient to entirely describe the second order statistics of a c.r.v. This is because it ignores the potential covariance between the parts of a variable and only describes their average variability. Another measure of variability of a complex random variable is the so called "pseudo-variance" [@reference], which can be obtained by applying the conventional formula of variance directly to the c.r.v. without the multiplication by conjugate:
\begin{equation}
    \begin{aligned}
    \varsigma_x^2 = & \mathrm{E}(x-\mu^2) = \mathrm{E}\left(((x_r-\mu_{r}) + i (x_i-\mu_{i}))^2\right) = \\
                    & \mathrm{E}((x_r-\mu_{r})^2) - \mathrm{E}((x_i-\mu_{i})^2) + i2 \mathrm{E}((x_r-\mu_{r})(x_i-\mu_{i}))
    \end{aligned}
    (\#eq:crvMomentSecondVariancePseudo)
\end{equation}
or using the notation $\sigma_{x_r,x_i}$ for covariance between the real and imaginary parts:
\begin{equation}
    \varsigma_x^2 = \sigma_{x_r}^2 - \sigma_{x_i}^2 + i2 \sigma_{x_r,x_i}.
    (\#eq:crvMomentSecondVariancePseudoShort)
\end{equation}
In the literature, the moment \@ref(eq:crvMomentSecondVariancePseudoShort) is called pseudo-variance [@reference], because it does not measure the variability of a variable, but rather gives a different information: it shows whether the real and imaginary variances are similar and what the covariance between them is. If both real and imaginary parts of \@ref(eq:crvMomentSecondVariancePseudoShort) are equal to zero then it is said that the distribution of the complex variable $x$ is spherical [@reference], i.e. the variances are similar and the real and imaginary parts are not linearly related. Furthermore, we should point out that while minimising \@ref(eq:crvMomentSecondVarianceShort) implies minimising both variances of the real and imaginary parts of $x$ (ignoring the covariance between them), minimising \@ref(eq:crvMomentSecondVariancePseudoShort) is not as straightforward. At very least, we could tell that it implies making the distribution of $x$ closer to the spherical one, but we cannot provide any thorough insight about this.

We should note that both measures can be considered as complex, and the main distinction between them is the multiplication by the same c.r.v or by its conjugate. As such we propose to call them respectively "conjugate variance" and "direct variance".

In R, the conjugate and direct variances are available in `cvar()` function of `complex` package.

```{r}
# Create a random variable
x <- complex(real=rnorm(100, mean=50, sd=5),
             imaginary=rnorm(100, mean=100, sd=10))
# Calculate the conjugate variance
cvar(x, method="conjugate") |>
    setNames("Conjugate variance")
# Calculate the direct variance
cvar(x, method="direct") |>
    setNames("Direct variance")
```

As we see from the output above, the direct variance has the negative real part, which indicates that the imaginary part has higher variance than the real one. The imaginary part of the direct variance shows the double covariance between the real and imaginary parts. Finally, the conjugate variance should be $5^2 + 10^2 = 125$, but due to small sample (100 observations in the generated random variable above), it will be equal to a number close to it.

Given that any complex variable can be represented in a vector form, the c.r.v. $x$ can be treated as a bivariate random variable, for which a covariance matrix can be calculated via:
\begin{equation}
    \boldsymbol{\Sigma}_x = \begin{pmatrix} \sigma_{x_r}^2 & \sigma_{x_r, x_i} \\ \sigma_{x_r, x_i} & \sigma_{x_i}^2 \end{pmatrix} .
    (\#eq:crvMomentSecondVarianceMatrix)
\end{equation}
The minimisation of the matrix \@ref(eq:crvMomentSecondVarianceMatrix) does not make sense, but instead it is possible to minimise the determinant of that matrix, which is called "Generalised Variance" (GV):
\begin{equation}
    \mathrm{GV} = |\boldsymbol{\Sigma}_x| = \sigma_{x_r}^2 \sigma_{x_i}^2 - \sigma_{x_r, x_i}^2 .
    (\#eq:crvMomentSecondGV)
\end{equation}

In R, this is implemented in `covar()` function from the `complex` package:
```{r}
covar(x)
```

The minimisation of GV, as can be seen from the formula \@ref(eq:crvMomentSecondGV), implies the simultaneous minimisation of variances of real and imaginary parts and maximisation of the square of covariance between them, making the resulting distribution compacter and emphasising the potential relations between the real and imaginary parts of a c.r.v. For our example in R, the GV equals to:

```{r}
covar(x) |> det()
```

All the three moments discussed in this subsection rely on variances of real and imaginary parts of a c.r.v. and on a covariance between those parts. These moments in turn can be calculated using the conventional formulae, correcting for the potential small sample bias [@referenceSBA]:
\begin{equation}
    \begin{aligned}
        \hat{\sigma}_{x_r}^2 = & \frac{1}{n-k} \sum_{j=1}^n (x_{r,j}-\bar{x}_{r,j})^2 \\
        \hat{\sigma}_{x_i}^2 = & \frac{1}{n-k} \sum_{j=1}^n (x_{i,j}-\bar{x}_{i,j})^2 \\
        \hat{\sigma}_{x_r, x_i} = & \frac{1}{n-k} \sum_{j=1}^n (x_{r,j}-\bar{x}_{r,j})(x_{i,j}-\bar{x}_{i,j})^2 ,
    \end{aligned}
    (\#eq:crvMomentSecondSample)
\end{equation}
where $k$ is the number of estimated parameters in a model.

Finally, similar to the variance it is possible to calculate second moments between two complex random variables $x = x_r+i x_i$ and $y = y_r + i y_i$ [@Picinbono]. The "conjugate" covariance is calculated similarly to \@ref(eq:crvMomentSecondVariance) via the multiplication by conjugate:
\begin{equation}
    \begin{aligned}
    \sigma_{x,y} = & \mathrm{E}((\tilde{x}-\tilde{\mu}_x) (y-\mu_y)) = \\
                   & \mathrm{E}\left(((x_r-\mu_{x,r}) - i (x_i-\mu_{x,i}))((y_r-\mu_{y,r}) + i (y_i-\mu_{y,i}))\right) = \\
                   & \mathrm{E}((x_r-\mu_{x,r})(y_r-\mu_{y,r})) + \mathrm{E}((x_i-\mu_{x,i})(y_i-\mu_{y,i})) + \\
                   & i \left(\mathrm{E}((x_r-\mu_{x,r})(y_i-\mu_{y,i})) - \mathrm{E}((x_i-\mu_{x,i})(y_r-\mu_{y,r}))\right)
    \end{aligned}
    (\#eq:crvMomentSecondCovariance)
\end{equation}
where $\mu_{x}$ and $\mu_y$ are the respective first moments of c.r.v. $x$ and $y$. Using the notations above, the same covariance can be rewritten as:
\begin{equation}
    \sigma_{x,y} = \sigma_{x_r, y_r} + \sigma_{x_i, y_i} + i (\sigma_{x_r, y_i} - \sigma_{x_i, y_r}),
    (\#eq:crvMomentSecondCovarianceShort)
\end{equation}
which similarly to the variance \@ref(eq:crvMomentSecondVarianceShort) mixes the moments of the real and imaginary parts of the two random variables. Note though that if the conjugate of $y$ is used instead of the conjugate of $x$, the imaginary part of the covariance \@ref(eq:crvMomentSecondCovarianceShort) will change to $\sigma_{x_i, y_r} - \sigma_{x_r, y_i}$, giving potentially different information about the relation between the variables.

In order to have more information about a c.r.v., we also need to consider the "direct" covariance [also known in the literature as pseudo-covariance, @reference], which can be shown to be equal to:
\begin{equation}
    \varsigma_{x,y} = \sigma_{x_r, y_r} - \sigma_{x_i, y_i} + i (\sigma_{x_i, y_r} + \sigma_{x_r, y_i}).
    (\#eq:crvMomentSecondPseudoCovarianceShort)
\end{equation}
Note that none of these moments on its own gives enough information about the relation between two complex variables, so they need to be used jointly. Alternatively, using vector representation, a covariance matrix between the two c.r.v. can be used to get a better understanding about the relations between them:
\begin{equation}
    \boldsymbol{\Sigma}_{x,y} =
        \begin{pmatrix}
            \sigma_{x_r}^2 & \sigma_{x_r, x_i} & \sigma_{x_r, y_r} & \sigma_{x_r, y_i} \\
            \sigma_{x_r, x_i} & \sigma_{x_i}^2 & \sigma_{x_i, y_r} & \sigma_{x_i, y_i} \\
            \sigma_{x_r, y_r} & \sigma_{x_i, y_r} & \sigma_{y_r}^2 & \sigma_{y_r, y_i} \\
            \sigma_{x_r, y_i} & \sigma_{x_i, y_i} & \sigma_{y_r, y_i} & \sigma_{y_i}^2
        \end{pmatrix} .
    (\#eq:crvMomentSecondCoVarianceMatrix)
\end{equation}

In R, the complex covariances are implemented in `ccov()` function from the `complex` package:

```{r}
# Create a random variable x
x <- complex(real=rnorm(100, mean=50, sd=5),
             imaginary=rnorm(100, mean=100, sd=10))
# Create a random variable y
y <- (1.5 + 3i) + (0.5 - 0.75i) * x +
            complex(real=rnorm(100, mean=0, sd=10),
                    imaginary=rnorm(100, mean=0, sd=10))
# Calculate the conjugate variance
ccov(x, y, method="conjugate") |>
    setNames("Conjugate covariance")
# Calculate the direct variance
ccov(x, y, method="direct") |>
    setNames("Direct covariance")
```

The functions `cvar()` and `ccov()` also accept a matrix instead of vector `x`, in which case they will produce a matrix of moments with complex variances on diagonal and complex covariances on the off-diagonals.

```{r}
ourData <- cbind(x,y)
cvar(ourData, method="direct")
```

While there exist higher order moments for complex random variables, we do not discuss them in this book.


## Parametric Distributions of c.r.v.
Similarly to the real valued random variables, complex random variables can follow some parametric distributions. In this section, we discuss several simple ones, some of which will be used later in this monograph.

In general, it is reasonable to assume that a c.r.v. has a distribution of a shape of a circle or ellipse on the complex plane. This becomes more apparent if we consider the exponential form of a c.r.v. \@ref(eq:complexNumberExponential) and consider both magnitude and the angle as random variables. In the simplest case, when they are not correlated and we consider a variable that is centred around the origin, the randomness coming from both of these components will result in a circle on the complex plane, see the following example in R (and Figure \@ref(fig:crvGeneratedCircle)):

```{r crvGeneratedCircle, fig.cap="C.r.v. generated from a uniform distribuiton.", fig.width=4, fig.height=4}
# Generate magnitudes from the uniform distribution
R <- runif(1000,0,10)
# Generate angles from the uniform distribution
phi <- runif(1000,-pi,pi)
# Create c.r.v in exponential form
x <- R * exp(complex(imaginary=phi))
# Plot the variable
plot(x)
```

The example in Figure \@ref(fig:crvGeneratedCircle) demonstrates how a c.r.v. with uncorrelated real and imaginary parts looks on the complex plane. If there is a relation between the parts then the circle transforms into ellipse. So, it is only logical to consider the distributions that rely on circle or ellipse for the c.r.v. instead of any other geometric shapes.

When it comes to distribution functions, in case of c.r.v. in general we should consider the joint bivariate distribution, thus all the probability density and cumulative distribution functions will be plotted in 3D with the density/probability in z-axis and the complex variable in x and y axes.


### Elliptic uniform distribution
The elliptic uniform distribution of c.r.v. is based on the equation of ellipse:
\begin{equation}
    \frac{(x_r-a_1)^2}{a_2^2} + \frac{(x_i-b_1)^2}{b_2^2} - \frac{(x_r-a_1) (x_i-b_1)}{a_2 b_2} = 1,
    (\#eq:ellipseFormula)
\end{equation}
where $a_1$ and $b_1$ are the shifts along the real and imaginary axes respectively, while $a_2$ and $b_2$ are the values that characterise the width and the height of the ellipse along the axes at the origin. The area of this ellipse is:
\begin{equation}
    A = \pi a_2 b_2 .
    (\#eq:EllipseArea)
\end{equation}
Based on this 
The PDF based on that is:
\begin{equation} \label{eq:EllipseArea2}
    f_z(z_j) = \left\{
        \begin{matrix}
            \frac{1}{\pi a b} & \text{when $z_j$ satisfies \eqref{eq:Ellipse}} \\
            0 & \text{otherwise}
        \end{matrix}
    \right. .
\end{equation}



### Normal distribution

+ Spherical c.r.v.


### Hotelling's distribution


